Verifying Hierarchical Cache Coherence Protocols: a theorem-prover based modular, reusable proof

### 2014-02-03 paper/Introduction.tex
Intuition

Because verified protocols are often described via Term Rewrite
Systems (TRS). It was shown by Dave
\etal{}~\cite{DNA:CoherenceImplementation} that a verified protocol
could be transliterated into Bluespec, a hardware description language
based on Guarded Atomic Actions, a restriction of
TRS~\cite{Hoe:TCAD}. This reduces the correctness of the
implementation to the correctness of the protocol, effectively
removing any implementation bugs. While the current Bluespec compiler
has certain restrictions which limit the expressiveness of certain
operations \eg{} fairness is not guaranteed, atomic actions are always
implemented in a single cycle, work has been done to remove these
restrictions~\cite{DNA:CoherenceImplementation, Karczmarek}.

@@

For this approach to be truly viable, we need not a single proof of
correctness of a single design, but a proof framework covering any
variation of a design that might be of interest for performance
reasons. This would enable implementation level changes without
compromising on the ability to verify correctness of the design.

>>>>

Because verified protocols are often described via Term Rewrite
Systems (TRS). It was shown by Dave
\etal{}~\cite{DNA:CoherenceImplementation} that such a verified
protocol could be transliterated into Bluespec, a hardware description
language based on Guarded Atomic Actions, a restriction of
TRS~\cite{Hoe:TCAD}. This reduces the correctness of the
implementation to the correctness of the protocol, effectively
removing any implementation bugs. While the current Bluespec compiler
has certain restrictions which limit the expressiveness of certain
operations \eg{} fairness is not guaranteed, atomic actions are always
implemented in a single cycle, work has been done to remove these
restrictions~\cite{DNA:CoherenceImplementation, Karczmarek}.

@@

For this approach to be truly viable, designers cannot be handed a
single proof of correctnessof a single design; rather, they need a
proof framework covering any variation of a design that might be of
interest for performance reasons. This would enable implementation
level changes without compromising on the ability to verify
correctness of the design.

### 2014-02-03 paper/Intuition.tex
some modifications of background
Intuition

\section{Intuition}

In this section we give the proof of the MSI cache coherence protocol. We give
the proof in a top-down fashion starting with the store atomicity theorem. We
specify important invariants and show how they are proved in the MSI protocol.
In Section \ref{Refinements} show how these invariants hold in other protocols
like MESI and MOSI. These invariants are true at each step in the (infinite)
list of transitions starting from the initial state. We use the term time $t$
to specify the global state reached after $t$ transitions.  All the invariants
below pertain to a particular address; we will implicitly assume a universal
quantification over all address as a prefix to each of the invariants (if not
explicitly mentioned).

\subsection{Top invariants}
\begin{defn}
\textit{clean}: A cache is said to be \clean{} for an address iff the state of the
cache is at least \Sh{} and the state of all its children is at most \Sh.
\end{defn}
\begin{inv}
At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
up-to-date value for that address, \ie
\begin{enumerate}
\item $c.\data[a] = \initData(a)$ and no store request has been processed at
any time $< t$, or
\item There is a store request $q$ that was processed at time $t'$ such that
$t' < t$ and $a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
request $q'$ was processed at any time $t_i$ such that $t' < t_i < t \wedge a =
q'.\addrQ$.
\end{enumerate}
\label{latestValue}
\end{inv}

@@

Using Invariants \ref{latestValue} to \ref{noReqAgain}, it is easy to see that
Theorem \textit{storeAtomicity} can be proved. Since a store can happen only
when the state of a leaf cache is in \Mo{} state (Invariant
\textit{processReq}), because of Invariant \textit{uniqM} only one store
request can be processed at a time throughout the system, thereby satisfying
\textit{uniqStTimes}. Because of Invariant \textit{noReqAgain}, once a response
has been produced for a request $i$, the cache can only process requests $j, j
> i$. Thus there is at most one response to each request, satisfying
\textit{uniqRespLabels}. A load request can be processed only if a cache is at
least in \Sh{} state according to Invariant \textit{processReq}.  When a leaf
cache is at least in \Sh{} state, since it has no children, it is \clean{}.
Thus, by Invariant \textit{lastestValue}, the final condition for Theorem
\textit{storeAtomicity} is satisfied.

Note that Invariants \textit{processReq} and \textit{noReqAgain} can be proved
directly from the atomic transitions describing the system (Figure \ref{trans}
for MSI; other protocols must also have similar transitions). Proving Invariants
\textit{uniqM} and \textit{latestValue} requires the following invariants.

\begin{inv}
\textit{conservative}: At all times, the state of a cache is not greater than
its parent's directory corresponding to the cache.
\label{conservative}
\end{inv}

@@

For any two caches $c$ and $p$, such that $c$ is a descendent of $p$ in the cache
hierarchy (the descendent relation between two caches is the transitive
reflexive closure of the parent relation), from Invariants
\textit{conservative} and \textit{stateCompat}, it is clear that for any
address $a$, $c.\state[a] \le p.\state[a]$. Any two caches $c_1, c_2$ in which
neither is the decendent of the other will have a lowest common ancestor $x$
such that both the caches are descendents of $x$ and $c_1, c_2$ are descendents
of distinct caches $x_1, x_2$, respectively, such that $x_1, x_2$ are children
of $x$. $x_1$ and $x_2$ are compatible because of Invariant
\textit{siblingCompat}. $c_1.\state[a] \le x_1.\state[a]$ and
$c_2.\state[a] \le x_2.\state[a]$ because of Invariants \textit{conservative}
and \textit{stateCompat}. From this, it can easily be seen from the definition
of compatible that $c_1$ and $c_2$ are compatible whenever $x_1$ and $x_2$ are
compatible. From the definition of compatible it can be seen that both caches can not
be in the \Mo{} state for them to be compatible. This directly proves Invariant
\textit{uniqM}.

Invariant \textit{latestValue} can be proved using induction on time. Let us
assume that all caches which are clean at time $t'$ such that $t'<t+1$ have the
latest value. If a cache $c$ is clean at time $t$, then none of its descendents
are in \Mo, and hence none of its descendents can process a store request at
$t$. Any non descendent is compatible with it (as shown in the previous
paragraph) and so no non-descendent cache can be in \Mo, since a clean cache is
at least in \Sh. So only $c$ can process a store request at $t$ (assuming $c$
is a leaf cache and is in state \Mo) when it is clean, and no other cache can
process a store request. By induction hypothesis, it contained the latest value
at time $t$ which remains true (because of no processing of store request) at
time $t+1$. Now, if a cache was not clean at time $t$, then it would have
received the data from some cache which was clean at the time of sending the
data (by Invariant \textit{noTransitWrite}), which means the data is the latest
value by induction hypothesis, so the new data it receives at time $t$ is the
latest value, making the clean cache at time $t+1$ have the latest value.

Invariant \ref{conservative} can be proved by
further breaking them into the following invariants between a parent-child pair
of caches.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A child sends a response to its parent iff it downgrades its state.
\item A parent sends a response to a child iff it upgrades the directory of
that child.
\item The state and the directory can change only on sending and receiving a
response; the receiving entity changes its value to that of the sending entity
immediately after sending the response message.
\item 
\end{enumerate}
\label{stateChange}
\end{inv}

@@

Invariant \textit{conservative} at any time $t$ for child cache $c$ and parent
$p$ for an address $a$ can be proved as follows. Consider the case when
$c.\state[a]$ never changed before $t$, then $c.\state[a] = \In$ because of the
initialization conditions described in Section \ref{System}. No response would
have been sent by $c$ to its parent for address $a$ before $t$, otherwise
$c.\state[a]$ would have changed. Thus the parent would not have received any
response from $c$, $p.\dstate[c][a]$ could have changed only by sending a
response to $c$ which can only upgrade it from the initial value of $\In$. Thus
at $t$, $p.\dstate[c][a] \ge c.state[a]$.

@@

>>>>

\section{Intuition}
\label{Sec:ProofStructure}

In this section, we structure our proof framework for our cache
coherence protocol. Specifically, we formally reduce the proof of the
whole system store atomicity theorem to the proof of a set of easily
proven invariants based on the structure of the protocol and design
structure which capture the ``intuitive'' proof of correctness of our
protocol design. As we show, these invariants are easily shown in the
base protocol design; further, in Section~\ref{Sec:Refinements}, we
show how the verification of a design refinement requires only that we
prove these invariants for the new system which is relatively
straightforward.

Our invariants hold before and after any execution of the component
transactions of our design reachable from the initial state. We use
the term time $t$ to specify the global state reached after $t$
transitions.  All the invariants below pertain to a particular
address; we will implicitly assume a universal quantification over all
address as a prefix to each of the invariants (if not explicitly
mentioned).

While we have proofs in Coq of each step in this framework our
discussion will be mainly informal for brevity. A full description can
be found at~\textbf{MURALI PUT REFERENCE HERE}.

\subsection{Reducing correctness to Node-level Invariants}

We've defined correctness is a whole system property. The first step
in our framework is to reduce the store atomicity property down to the
following 4 global invariants about nodes:

\begin{defn}
\textit{clean}: A cache is said to be \clean{} for an address $a$ iff
the state of the cache for that address is at least \Sh{} and the
state of all its children is at most \Sh.
\end{defn}
\begin{inv}
At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
up-to-date value for that address, \ie{}
\begin{enumerate}

\item $c.\data[a] = \initData(a)$ and no store request has been
  processed at any time $< t$, or

\item There is a store request $q$ that was processed at time $t'$ such that
$t' < t$ and $a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
request $q'$ was processed at any time $t_i$ such that $t' < t_i < t \wedge a =
q'.\addrQ$.
\end{enumerate}
\label{latestValue}
\end{inv}

@@

We can see that the correctness of our system
(Theorem~\textit{storeAtomicity}) follows naturally from Invariants
\ref{latestValue} to \ref{noReqAgain}. Since a store may happen only
when leaf cache is in \Mo{} state (Invariant~\textit{processReq}), and
only one cache may cause a store request per address at a particular
point in time (Invariant \textit{uniqM}), Theorem~\textit{uniqStTimes}
holds. Given Invariant~\textit{noReqAgain}, once a response has been
produced for a request $i$, the cache may only process later requests
$j, j > i$. Thus there is at most one response to each request,
satisfying Theorem~\textit{uniqRespLabels}. A load request can be
processed only if a cache is at least in \Sh{} state according to
Invariant \textit{processReq}.  When a leaf cache is at least in \Sh{}
state, since it has no children, it is \clean{}.  Thus, by
Invariant~\textit{lastestValue}, the final condition for
Theorem~\textit{storeAtomicity} is satisfied.

\subsection{Reducing to Pair-wise Invariants}

Invariants~\textit{processReq} and \textit{noReqAgain} can be proved
directly from the atomic transitions describing the system
(Figure~\ref{trans} for MSI; other protocols must also have similar
transitions). Proving invariants~\textit{uniqM} and
\textit{latestValue} needs some knowledge of the hierarchy. We can
prove both using the following invariants on parent-child
relationships.

\begin{inv}
\textit{conservative}: At all times, the state of an address in a
cache is not greater than its that in the corresponding entry in the
parent's directory.
\label{conservative}
\end{inv}

@@

For any two caches $c$ and $p$, such that $c$ is a descendent of $p$ in the cache
hierarchy (the descendant relation between two caches is the transitive
reflexive closure of the parent relation), from Invariants
\textit{conservative} and \textit{stateCompat}, it is clear that for any
address $a$, $c.\state[a] \le p.\state[a]$. Any two caches $c_1, c_2$ in which
neither is the decendant of the other will have a lowest common ancestor $x$
such that both the caches are descendants of $x$ and $c_1, c_2$ are descendants
of distinct caches $x_1, x_2$, respectively, such that $x_1, x_2$ are children
of $x$. $x_1$ and $x_2$ are compatible because of Invariant
\textit{siblingCompat}. $c_1.\state[a] \le x_1.\state[a]$ and
$c_2.\state[a] \le x_2.\state[a]$ because of Invariants \textit{conservative}
and \textit{stateCompat}. From this, it can easily be seen from the definition
of compatible that $c_1$ and $c_2$ are compatible whenever $x_1$ and $x_2$ are
compatible. From the definition of compatible it can be seen that both caches can not
be in the \Mo{} state for them to be compatible. This directly proves Invariant
\textit{uniqM}.

Invariant~\textit{latestValue} can be proved using induction on time. Let us
assume that all caches which are clean at time $t'$ such that $t'<t+1$ have the
latest value. If a cache $c$ is clean at time $t$, then none of its descendents
are in \Mo, and hence none of its descendents can process a store request at
$t$. Any non descendent is compatible with it (as shown in the previous
paragraph) and so no non-descendent cache can be in \Mo, since a clean cache is
at least in \Sh. So only $c$ can process a store request at $t$ (assuming $c$
is a leaf cache and is in state \Mo) when it is clean, and no other cache can
process a store request. By induction hypothesis, it contained the latest value
at time $t$ which remains true (because of no processing of store request) at
time $t+1$. If a cache was not clean at time $t$, then it would have
received the data from some cache which was clean at the time of sending the
data (by Invariant~\textit{noTransitWrite}), which means the data is the latest
value by induction hypothesis, so the new data it receives at time $t$ is the
latest value, making the clean cache at time $t+1$ have the latest value.

Invariant~\textit{conservative} can be proved by
further breaking them into the following invariants between a parent-child pair
of caches.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A child sends a response to its parent iff it downgrades its state.
\item A parent sends a response to a child iff it upgrades the directory of
that child.
\item The state and the directory can change only on sending and receiving a
response; the receiving entity changes its value to that of the sending entity
immediately after sending the response message.
\end{enumerate}
\label{stateChange}
\end{inv}

@@

Invariant~\textit{conservative} at any time $t$ for child cache $c$ and parent
$p$ for an address $a$ can be proved as follows. Consider the case when
$c.\state[a]$ never changed before $t$, then $c.\state[a] = \In$ because of the
initialization conditions described in Section \ref{System}. No response would
have been sent by $c$ to its parent for address $a$ before $t$, otherwise
$c.\state[a]$ would have changed. Thus the parent would not have received any
response from $c$, $p.\dstate[c][a]$ could have changed only by sending a
response to $c$ which can only upgrade it from the initial value of $\In$. Thus
at $t$, $p.\dstate[c][a] \ge c.state[a]$.

@@

\subsection{Final reduction}

### 2014-02-03 paper/System.tex
Intuition

The behavior of the system are described via atomic state transitions.
These transitions have two parts: the guard dictating when the
transition can fire, and the action defining how the state of the
system changes when the transition happens. These transitions or \emph{guarded
atomic actions} logically happen one by one, making the overall behavior of the
system well defined. Figure~\ref{trans} gives the atomic transitions for a
cache coherent memory subsystem. The actions in the transitions represent only
the specific system states that change. We use $x \Leftarrow y$ to indicate
setting of a specific state $x$ of the overall system to value $y$ because of a
transition.

>>>>

The behavior of the system are described via the atomic state
transitions.  These transitions have two parts: the guard dictating
when the transition can fire, and the action defining how the state of
the system changes when the transition happens. In our generated
implementation, these transitions or \emph{guarded atomic actions}
happen concurrently in a single hardware clock, but they will
logically happen happen one at a time. Figure~\ref{trans} gives the
atomic transitions for a cache coherent memory subsystem. The actions
in the transitions represent only the specific system states that
change. We use $x \Leftarrow y$ to indicate setting of a specific
state $x$ of the overall system to value $y$ because of a transition.
